Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-23 paper/main.tex
pass back to Steve

\subsection{Bayesian Inference}

@@

We then fit our DGP.FCO model on each of 111 training cosmologies, reserving 6 cosmologies
for later hold-out testing.  To speed-up burn-in of MCMC chains, we conduct
50,000 MCMC iterations for the first cosmology, then initialize the chains for all 
other cosmologies with these burned-in samples, so fewer iterations are needed for the
remaining 110 cosmologies.  Ultimately, we conduct 10,000 iterations for each of these,
removing 5,000 for burn-in, then thinning by 5.  Figure \ref{fig:plot_fit} 
visualizes the DGP.FCO fit for the first training cosmology.  We center the plot
by subtracting the posterior mean in order to more easily visualize the credible
interval bounds.  The UQ from the DGP.FCO model accounts for varying precisions 
across differing $k$ values, as well as the spatial correlation within the 
functional realizations.  To provide insight into the nonstationarity of the power
matter spectrum, the right panel of Figure \ref{fig:plot_fit} shows the elliptical slice samples of $W$.
These warpings depart from the identiy mapping around $x\approx-1$, 
stretching inputs where dynamics are shifting more rapidly for larger $k$ values.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.jpeg}
    \caption{{\it Left:} 95\% credible intervals for the power spectrum of the first Mira-Titan
    cosmology (centered by the predicted posterior mean). Perturbation theory, 
    low-resolution, and high-resolution runs are shown for context. {\it Right:} Posterior draws
    of $W$, illustrating some nonstationary behavior.}
    \label{fig:plot_fit}
\end{figure}

@@

Our goal is to predict the power spectrum for a held-out cosmology. 
For simplicity, we will denote these eight parameters for a given cosmology (see 
Section \ref{sec:data}) as $\psi \in \mathbb{R}^{p_\psi}$. With our estimated posterior means
from DGP.FCO for each training cosmology, we now predict $S$ for unobserved cosmologies
as a function of $\psi$.

We have $m$-many training observations ($j = 1,\dots,m$) which each contain the $p_\psi$-dimensional
input $\psi_j$ and DGP.FCO estimates $\hat{S}_j$ for the $j$th cosmology. 
We combine all $\hat{S}_j$ estimates into a $n\times m$ matrix $\boldsymbol\eta$.
Due to the fact that each column of $\boldsymbol\eta$ represents functional output, 
we choose to reformulate the problem using principal components (PCs). Details of this can be found in 
\ref{sec:apdx_basis}. Here we will use the first $p_\eta$ PCs.

Figure \ref{fig:mean_PCs_oneW} illustrates this decomposition with the estimated 
mean trend for the Mira-Titan dataset, the $p_\eta$ basis functions, and one 
example of how the weight for the first PC will vary dependent on the first of 
$p_\psi=8$ cosmological parameters contained in $\psi$. As a result, our goal becomes 
estimating the weights $\Gamma$ corresponding to each of the $p_\eta$ PCs, 
as a function of $\psi$.

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model the principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. That is, the $i$th vector of PC weights $w_i(\psi)$, 
$i \in \{1,\ldots,p_\eta\}$, is modeled as a zero-mean Gaussian process with a 
squared exponential (i.e., Gaussian) covariance function:
\[
w_i(\psi) \sim GP\left(0,\ \frac{1}{\lambda_{z_i}} R^{(i)} + \frac{1}{\lambda_{s_i}} I \right),
\]
with covariance function
\[
R^{(i)}_{uv} = \exp\left(-\sum_{k=1}^{p_\psi} \beta_{ik} (x_{uk} - x_{vk})^2 \right),
\]
where $\beta_{ik}$ is the inverse squared lengthscale for the $k$th input dimension, 
$\lambda_{z_i}$ is the marginal precision for $w_i(\psi)$, and $\lambda_{s_i}$ is the nugget 
precision (set to zero if no nugget is used).

Using these trained GPs from the PCs, we predict the spectra for each of the $N$ held-out 
cosmologies for both the CAMB and Mira-Titan datasets.

\subsection{Predicting Spectra for CAMB Data}
\label{subsec:camb_pred}

From our $m=32$ posterior means using DGP.FCO, these are input into SEPIA in order to predict power spectra
for the $N=32$ test cosmologies. In the CAMB dataset, $p_\psi=6$. With the infinite-resolution spectra, we 
have a sense of the ``best-case" predictions we can have (i.e., if we trained with the 32 true 
infinite-resolution spectra, as opposed to our estimated posterior means). In Figure \ref{fig:pca_preds_v_camb}, 
we see comparisons of the heldout infinite resolution alongside two predictions: one 
based on DGP.FCO and the ``best-case" scenario using the infinite-resolution spectra of 
the training set. There is a high level of agreement between the pairs of predictions 
across cosmologies. Upon comparing MSEs for each of these prediction scenarios 
(Figure \ref{fig:mse_camb}, most are very similar across the two groups (with the two most difficult 
predictions being the exceptions), suggesting that overall DGP.FCO is estimating the true spectra 
for each batch well.

\begin{figure}
    \centering
    \includegraphics[width=4in]{preds_diff1_4_8.jpg}
    \caption{Predictions for 16 of the $N=32$ test cosmologies after centering and scaling. 
             The interpolated infinite resolution run is shown in black, the prediction 
             (trained on the infinite resolution) is in blue and the prediction from 
             the DGP.FCO posterior mean is in red.}   
    \label{fig:pca_preds_v_camb}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=3in]{mse_dot.jpg}
    \caption{Comparison of MSE for CAMB predictions, when training with DGP.FCO posterior 
             means (left), and training with the ``true" infinite-resolution spectra (right).}   
    \label{fig:mse_camb}
\end{figure}

\subsection{Predicting Spectra from Mira-Titan Data}
\label{subsec:mira_pred}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{{\it Left:} The estimated mean trend of the $m=111$ posterior means from 
    the Mira-Titan dataset. {\it Middle:} The $p_\eta$ principal components obtained from 
    the posterior means. {\it Right:} An 
    illustration of how the weights for the first PC will change as the first 
    cosmological parameter varies \textcolor{red}{This last plot is wonky because 
    it was made with only the $N=6$ test cosmologies, not a large uniform or FF design, 
    Steve can update this}.}
    \label{fig:mean_PCs_oneW}
\end{figure}

Here we show prediction results for $N=6$ held-out cosmologies for the Mira-Titan dataset. 
We compare our method with Cosmic Emu \citep{moran2023mira}, the emulator constructed using the full suite 
of $m=111$ Mira-Titan model runs. Cosmic Emu uses a Bayesian approach where the smooth 
spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep{moran2024dpc}.

Both Cosmic Emu and DGP.FCO are capable of modeling a true underlying functions based 
upon multiple functional realizations, but our approach utilizes DGPs within a 
hierarchical model in lieu of deep process convolutions. After obtaining the estimated 
smooth spectra from the training cosmologies, each method utilizes GPs in order to 
estimate the weights of the PCs from the estimated spectra. Below, we compare prediction 
performance between Cosmic Emu and DGP.FCO for the $N=6$ held-out cosmologies. 
% \textcolor{magenta}{In addition, we can also predict over a large uniform or full 
% factorial design to obtain estimated main effects for each cosmological parameter in 
% $\psi$, interactions amongst the $p_\psi=8$ parameters, and decompose the overall 
% variation by main effects, two-way interactions, and so on. Should this be included?}

Plots for the average root mean squared error (RMSE) for each prediction method over the 
$N=6$ cosmologies and each $k$ value are shown in Figure \ref{fig:plot_rmse_k}. Comparisons 
of the spectrum estimation for training cosmologies are shown as well (using RMSE against 
the weighted average of observations). For predictions, DGP.FCO achieves the greatest 
reduction in RMSE in the region where perturbation theory and the low-resolution runs 
are deemed unbiased. For the region where only the high-resolution run is unbiased, 
Cosmic Emu has a lower average RMSE. Over all $k$ values and predictions, DGP.FCO has 
an average RMSE of 0.0089 and Cosmic Emu has an average RMSE of 0.0106. Our RMSE is 
lower at 56\% of $k$ values considered. The individual predictions for each method and 
cosmology are shown in Figure \ref{fig:plot_pred_1to6}.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{rmse_by_k.png}
    \caption{{\it Left:} Average RMSE of each prediction method across all $k$ values. 
             {\it Right:} Average RMSE for each method when using the held-out computer 
             model runs to train the model. In each case, the truth is assumed to 
             be the weighted average of the computer model runs.}
    \label{fig:plot_rmse_k}
\end{figure}

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{sims_MSE.jpeg}
    \caption{Boxplot of MSEs across two different functions and covariance settings 
             (lower is better). For each case, the boxplot is constructed from 20 
             random batches were simulated. Each column represents a function/variance 
             specification pair, and each row shows results for batch sizes of 
             either 5 (top) or 15 (bottom).}    
    \label{fig:sims_MSE}
\end{figure}

@@

>>>>

\subsection{Bayesian Inference}\label{sec:inference}

@@

We then fit our DGP.FCO model on each of 111 training cosmologies, reserving 6 cosmologies
for later hold-out testing.  Figure \ref{fig:plot_fit} 
visualizes the DGP.FCO fit for the first training cosmology.  We center the plot
by subtracting the posterior mean in order to more easily visualize the credible
interval bounds.  The UQ from the DGP.FCO model accounts for varying precisions 
across differing $k$ values, as well as the spatial correlation within the 
functional realizations.  To provide insight into the nonstationarity of the power
matter spectrum, the right panel of Figure \ref{fig:plot_fit} shows burned-in elliptical 
slice samples of $W$.
These warpings depart from the identiy mapping around $x\approx-1$, 
stretching inputs where dynamics are shifting more rapidly for larger $k$ values.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.jpeg}
    \caption{{\it Left:} 95\% credible intervals for the power spectrum of the first Mira-Titan
    cosmology (centered by the predicted posterior mean). Perturbation theory, 
    low-resolution, and high-resolution runs are shown for context. {\it Right:} Posterior draws
    of $W$, illustrating nonstationary behavior by departing from a linear (identity) mapping.}
    \label{fig:plot_fit}
\end{figure}

@@

Let $\psi \in \mathbb{R}^{p_\psi}$ denote the set of parameters that define a particular cosmology.
Mira-Titan has 8 such parameters as described in Section \ref{sec:data}; CAMB has 6.
Our goal is to effectively predict the power matter spectrum as a function of $\psi$, i.e. $\hat{S}(\psi)$, 
for a held-out cosmology.  We observe $m$-many training cosmologies and fit our DGP.FCO 
model (Section \ref{sec:inference}) to each one.  Let $\hat{S}_j$ for $j \in \{1,\ldots,m\}$ denote
the estimated posterior mean of each training cosmology.  In total, our training data comprises
the $p_\psi$-dimensional input $\psi_j$ and the $n$-dimensional output $\hat{S}_j$ (which is a 
function of $x_i = \log_{10}(k_i)$) for $j \in \{1,\ldots,m\}$.

To handle the functional nature of this output, we reformulate the problem using principal
components (PCs; \textcolor{red}{@Steve - can you add a textbook reference?}).  
We first center all $\hat{S}_j$ by subtracting $\bar{S} = \frac{1}{111}\sum_{j=1}^{111} \hat{S}_j$, then
we combine them into an $n\times m$ matrix $\boldsymbol\eta$.  We use principal components to decompose 
$\boldsymbol\eta = BW^\top$ where $B$ is the $n\times p_\eta$ matrix of basis functions and $W$
is the $m\times p_\eta$ matrix of basis weights.  Details on this decomposition are 
reserved for \ref{sec:apdx_basis}.  For the CAMB dataset, $p_\eta=6$; for Mira-Titan, $p_\eta=8$.
As a visual, the left panel of Figure \ref{fig:mean_PCs_oneW} shows $\bar{S}$ for the Mira-Titan 
training cosmologies.  The center panel shows the $p_\eta=8$ basis functions.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{{\it Left:} The estimated mean trend of the $m=111$ posterior means from 
    the Mira-Titan dataset. {\it Middle:} The $p_\eta$ principal components obtained from 
    the posterior means. {\it Right:} An 
    illustration of how the weights for the first PC will change as the first 
    cosmological parameter varies \textcolor{red}{This last plot is wonky because 
    it was made with only the $N=6$ test cosmologies, not a large uniform or FF design, 
    Steve can update this}.
    \textcolor{red}{This figure also needs work.  Axes need to be fixed.  Caption needs re-write.
    Right panel needs to be redone.  It should look smooth like a GP.  Or else just remove it.}}
    \label{fig:mean_PCs_oneW}
\end{figure}

Let $w_i^{(j)}$ denote the $ij^\textrm{th}$ element of $W$, representing the basis weight for the 
$j^\textrm{th}$ cosmology and the $i^\textrm{th}$ PC.
Conditional on $B$, our training data now consists of scalars $w_i^{(j)}$ as a function of $\psi_j$
for $j \in \{1,\ldots,m\}$ and $i \in \{1,\ldots,p_\eta\}$.  If we can effectively predict these basis weights
as a function of $\psi$ (i.e., $\hat{w}_i(\psi)$ for $i \in \{1,\ldots,p_\eta\}$), then we may obtain predictions 
of the power spectrum as
\[
\hat{S}(\psi) = \sum_{i=1}^{p_\eta} \hat{w}_i(\psi) B_i,
\]
where $B_i$ is the $i^\textrm{th}$ basis function (column) of $B$.

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model these principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. That is, each $w_i(\psi)$ is modeled as a 
zero-mean Gaussian process with a squared exponential (i.e., Gaussian) covariance function:
\[
w_i(\psi) \sim GP\left(0,\ \frac{1}{\lambda_{z_i}} R^{(i)} + \frac{1}{\lambda_{s_i}} I \right),
\]
with covariance function
\[
R^{(i)}_{uv} = \exp\left(-\sum_{k=1}^{p_\psi} \beta_{ik} (x_{uk} - x_{vk})^2 \right),
\]
where $\beta_{ik}$ is the inverse squared lengthscale for the $k$th input dimension, 
$\lambda_{z_i}$ is the marginal precision for $w_i(\psi)$, and $\lambda_{s_i}$ is the nugget 
precision (set to zero if no nugget is used).
\textcolor{red}{(@Steve - there is a lot missing here.  The inputs to the covariance kernel
here need to be $\psi$.  Are we using a nugget?  If not, just remove it.  How are these parameters
being estimated?  Make sure this kernel notation matches what we used earlier as neatly as possible.)}

Using these trained GPs from the PCs, we predict the spectra for each of the $N$ held-out 
cosmologies for both the CAMB and Mira-Titan datasets. \textcolor{red}{(@Steve - this needs work too.
Please include the formula for the GP posterior mean, which is how we obtain $\hat{w}_i$.  We never
had use to include it earlier, but we should use it here.  If you need, you can introduce new matrix/vector
notation to represent the training data in this scenario.  At this point, you should refer back to the right
panel of Figure \ref{fig:mean_PCs_oneW} and narrate how this shows the predicted GP mean for $w_1$ (for Mira-Titan)
as a function of the first cosmological parameter $\psi_1$.)}

\subsection{Predicting Spectra for CAMB}
\label{subsec:camb_pred}

\textcolor{red}{(@Steve - we removed the text earlier that talked about the training/testing split in the
wrong section.  You need to introduce it here now.)} From our $m=32$ posterior means using DGP.FCO, 
these are input into SEPIA \textcolor{red}{(What does this mean?  Just refer to the equations from the previous
section for how we obtain predictions.)} in order to predict power spectra
for the $N=32$ test cosmologies.

Since CAMB offers the ``true'' infinite-resolution spectra, we have the opportunity to benchmark our
results in two ways.  First, we may compare our predictions to the true power spectrum for each held-out
cosmology.  Second, we may leverage the infinite-resolution spectra for the training cosmologies in
place of our estimated $\hat{S}_j$, to get a sense of the ``best-case" predictions that could originate
from the PC decomposition.  Figure \ref{fig:pca_preds_v_camb} compares our predictions (DGP.FCO + PC, red)
to this ``best-case'' prediction (true + PC, blue) and the actual true spectrum for \textcolor{red}{10?} 
of the held-out test cosmologies.  There is a high level of agreement between the pairs of predictions 
across cosmologies.  We then compare the mean squared errors of our predictions and the ``best-case'' 
predictions across all held-out cosmologies, with results shown in Figure \ref{fig:mse_camb}.  We again
see only minimal differences, which we take to indicate that DGP.FCO is offering effective
estimates of the true spectra for each cosmology.

\begin{figure}
    \centering
    \includegraphics[width=4in]{preds_diff1_4_8.jpg}
    \caption{Predictions of power matter spectra for 16 of the CAMB test cosmologies,
             after centering and scaling. 
             The true infinite resolution run is shown in black, the prediction 
             (trained on the infinite resolution) is in blue, and the prediction from 
             the DGP.FCO posterior mean is in red.
             \textcolor{red}{(This figure does not look polished.  Make it wider.  I recommend
             showing 2 rows of 5 plots.  Also, change line types so we are less reliant on color.
             Missing x-axis label.  No need to number the panels.)}}   
    \label{fig:pca_preds_v_camb}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=3in]{mse_dot.jpg}
    \caption{Comparison of MSE for CAMB predictions, when training with DGP.FCO posterior 
             means (left), and training with the ``true" infinite-resolution spectra (right).
             \textcolor{red}{(This figure also needs refinement.)}}   
    \label{fig:mse_camb}
\end{figure}

\subsection{Predicting Spectra for Mira-Titan}
\label{subsec:mira_pred}

Here we show prediction results for $N=6$ held-out cosmologies for the Mira-Titan dataset. 
\textcolor{red}{(This sentence is misleading.... we only offer one figure that shows RMSE.)}
We compare our method with Cosmic Emu \citep{moran2023mira}, the state-of-the-art emulator 
constructed on the same $m=111$ Mira-Titan cosmologies. Cosmic Emu uses a Bayesian approach 
where the ``true'' spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep{moran2024dpc}.  \textcolor{red}{(Add sentence about hwo Cosmic Emu also uses a PC
approach to predict power matter spectra for held-out cosmologies as a function of $\psi$.)}

%Both Cosmic Emu and DGP.FCO are capable of modeling a true underlying functions based 
%upon multiple functional realizations, but our approach utilizes DGPs within a 
%hierarchical model in lieu of deep process convolutions. After obtaining the estimated 
%smooth spectra from the training cosmologies, each method utilizes GPs in order to 
%estimate the weights of the PCs from the estimated spectra. Below, we compare prediction 
%performance between Cosmic Emu and DGP.FCO for the $N=6$ held-out cosmologies. 
% \textcolor{magenta}{In addition, we can also predict over a large uniform or full 
% factorial design to obtain estimated main effects for each cosmological parameter in 
% $\psi$, interactions amongst the $p_\psi=8$ parameters, and decompose the overall 
% variation by main effects, two-way interactions, and so on. Should this be included?}

Using our proposed approach and Cosmic Emu, we obtain two sets of predictions for $\hat{S}_t(\psi_t)$ 
with $t\in\{1,\ldots,6\}$ indexing the six held-out cosmologies.  To summarize and compare
performance, we evaluate the root mean squared error across $t$ for each wavenumber $x_i = \log_{10}(k_i)$.
\textcolor{red}{(This is again missing key details.  What is the truth you are using to calculate
RMSE???)}  The left panel of Figure \ref{fig:plot_rmse_k} shows these RMSEs for the held-out cosmologies.
Comparisons of the spectrum estimation for training cosmologies are shown as well (using RMSE against 
the weighted average of observations). \textcolor{red}{(What does this previous sentence mean??? We
have discussed that the weighted average is not a good estimate of the true spectrum.  So we should not
be comparing our estimated means to the weighted averages.  I am so confused....)}

For predictions, DGP.FCO achieves the greatest 
reduction in RMSE in the region where perturbation theory and the low-resolution runs 
are deemed unbiased. For the region where only the high-resolution run is unbiased, 
Cosmic Emu has a lower average RMSE. Over all $k$ values and predictions, DGP.FCO has 
an average RMSE of 0.0089 and Cosmic Emu has an average RMSE of 0.0106. Our RMSE is 
lower at 56\% of $k$ values considered. The individual predictions for each method and 
cosmology are shown in Figure \ref{fig:plot_pred_1to6}.  \textcolor{red}{(Why is this figure in the appendix?
But even more concerning, why are we subtracting the weighted average?  In every other plot we subtract our
estimated mean.  This switch to using the weighted average in the final subsection is very concerning
to me.  Also - what is the reader supposed to be gleaning from each of these plots?  Any time you
have a figure, you need to include narration in the text that explains what the takeaways are.)}

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{rmse_by_k.png}
    \caption{{\it Left:} Average RMSE of each prediction method across all $k$ values. 
             {\it Right:} Average RMSE for each method when using the held-out computer 
             model runs to train the model. In each case, the truth is assumed to 
             be the weighted average of the computer model runs.
             \textcolor{red}{(Shouldn't the average be across the 6 cosmologies?  Saying it 
             is across the $k$ values is misleading.  Again, I am concerned for us to use
             the weighted average as the truth.  This seems to go starkly against things
             that we have said earlier in the paper.)}}
    \label{fig:plot_rmse_k}
\end{figure}

@@

\textcolor{red}{(@Steve - I think you should revisit this section once the rest of the
revisions are done.  Make sure it aligns with what the paper ends up being.  I am not going
through this sectio myself at this time.)}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{sims_MSE.jpeg}
    \caption{Boxplot of MSEs across two different functions and covariance settings 
             (lower is better). For each case, the boxplot is constructed from 20 
             random batches were simulated. Each column represents a function/variance 
             specification pair, and each row shows results for batch sizes of 
             either 5 (top) or 15 (bottom).}    
    \label{fig:sims_MSE}
\end{figure}

@@

\textcolor{red}{(@Steve - revise the text so it works for any $m$, not just 111.  
Also, replace $\Gamma$ with $W$ so it matches the text I revised earlier.)}
